---
title: "Logistic GEM Simulations"
author: "Clay Cressler"
date: "24/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I want to simulate the logistic GEM with different parameter sets to explore how the variation in parameters impacts the location of the eco-evolutionary equilibrium and the TEA that is reached by the evolving population.

```{r, echo=FALSE}
source("logistic_GEM.R")
library(parallel)
library(tidyverse)
```
```{r}
set.seed(123947)
seeds <- floor(runif(200, 1, 1e6))
mclapply(seeds,
           function(s) logistic_GEM(seed=s, dt=0.1, tmax=600, N0=5, traitmean=1.8, traitcv=0.3, h2=0.75, bs=0.075, ds=0.075, slope=0.3/1.8^2), 
           mc.cores=5) -> out
out0 <- out
## for illustrative purposes, just work with the reps that didn't go extinct
out[((lapply(out, function(x) max(x[[1]])) %>% unlist) > 590) %>% which] -> out

lapply(1:length(out), 
         function(i) data.frame(time=out[[i]][[1]], N=lapply(out[[i]][[2]], length) %>% unlist, 
                                mean=lapply(out[[i]][[2]], mean) %>% unlist, 
                                var=lapply(out[[i]][[2]], var) %>% unlist, 
                                rep=i)
        ) %>% do.call(rbind.data.frame, .) -> allreps
mutate(allreps, time=floor(10*time)/10) %>% 
  group_by(time) %>% 
  summarise(N=mean(N), b=mean(mean)) -> meandata

## Eco-evo equilibrium
ecoevo <- c(5.4, 2.7/0.15)

N <- seq(2, 20, 0.1)
plot(N, 0.075*N+sqrt((0.075*N)^2+0.075*N/(0.3/1.8^2)), type='l', lwd=2, col=grey(0.6), xlab="Population size", ylab=expression(b[max]))
points(18, 5.4, pch=21, cex=2, col=4, bg=4)
with(meandata, lines(N, b, lwd=2, col="orange"))



```

Start a bit closer to a TEA:

```{r}
set.seed(123947)
seeds <- floor(runif(200, 1, 1e6))
mclapply(seeds,
           function(s) logistic_GEM(seed=s, dt=0.1, tmax=400, N0=5, traitmean=2.4, traitcv=0.3, h2=0.75, bs=0.075, ds=0.075, slope=0.3/1.8^2), 
           mc.cores=5) -> out
## for illustrative purposes, just work with the reps that didn't go extinct
out[((lapply(out, function(x) max(x[[1]])) %>% unlist) > 390) %>% which] -> out

lapply(1:length(out), 
         function(i) data.frame(time=out[[i]][[1]], N=lapply(out[[i]][[2]], length) %>% unlist, 
                                mean=lapply(out[[i]][[2]], mean) %>% unlist, 
                                var=lapply(out[[i]][[2]], var) %>% unlist, 
                                rep=i)
        ) %>% do.call(rbind.data.frame, .) -> allreps
mutate(allreps, time=floor(10*time)/10) %>% 
  group_by(time) %>% 
  summarise(N=mean(N), b=mean(mean)) -> meandata

N <- seq(2, 20, 0.1)
plot(N, 0.075*N+sqrt((0.075*N)^2+0.075*N/(0.3/1.8^2)), type='l', lwd=2, col=grey(0.6), xlab="Population size", ylab=expression(b[max]))
points(18, 5.4, pch=21, cex=2, col=4, bg=4)
with(meandata, lines(N, b, lwd=2, col="orange"))


```

```{r}
set.seed(123947)
seeds <- floor(runif(1000, 1, 1e6))
mclapply(seeds,
           function(s) logistic_GEM(seed=s, dt=0.1, tmax=400, N0=5, traitmean=3.5, traitcv=0.3, h2=0.75, bs=0.075, ds=0.075, slope=0.3/1.8^2), 
           mc.cores=5) -> out
## for illustrative purposes, just work with the reps that didn't go extinct
out[((lapply(out, function(x) max(x[[1]])) %>% unlist) > 390) %>% which] -> out

lapply(1:length(out), 
         function(i) data.frame(time=out[[i]][[1]], N=lapply(out[[i]][[2]], length) %>% unlist, 
                                mean=lapply(out[[i]][[2]], mean) %>% unlist, 
                                var=lapply(out[[i]][[2]], var) %>% unlist, 
                                rep=i)
        ) %>% do.call(rbind.data.frame, .) -> allreps
mutate(allreps, time=floor(10*time)/10) %>% 
  group_by(time) %>% 
  summarise(N=mean(N), b=mean(mean)) -> meandata

N <- seq(2, 20, 0.1)
plot(N, 0.075*N+sqrt((0.075*N)^2+0.075*N/(0.3/1.8^2)), type='l', lwd=2, col=grey(0.6), xlab="Population size", ylab=expression(b[max]))
points(18, 5.4, pch=21, cex=2, col=4, bg=4)
with(meandata, lines(N, b, lwd=2, col="orange"))


```


```{r}

set.seed(123947)
seeds <- floor(runif(50, 1, 1e6))

cvals = seq(0.04, 0.14, 0.02)

for (j in 1:length(cvals)) {
  print(cvals[j])
  ecoevoequil <- c( 1/(4*cvals[j]*0.02), 1/(2*cvals[j]) )
  mclapply(seeds,
           function(s) logistic_GEM(seed=s, dt=0.1, tmax=400, N0=1/10*ecoevoequil[1], traitmean=1/5*ecoevoequil[2], traitcv=0.3, h2=0.8, bs=0.01, ds=0.01, slope=cvals[j]), 
           mc.cores=5) -> out
  lapply(1:length(out), 
         function(i) data.frame(time=out[[i]][[1]], N=lapply(out[[i]][[2]], length) %>% unlist, 
                                mean=lapply(out[[i]][[2]], mean) %>% unlist, 
                                var=lapply(out[[i]][[2]], var) %>% unlist, 
                                rep=i)
        ) %>% do.call(rbind.data.frame, .) -> allreps
  mutate(allreps,
       time=floor(10*time)/10) %>% 
  group_by(time) %>%
  summarise(N=mean(N), mean=mean(mean), var=mean(var)) %>% 
  as.data.frame -> meandata
  assign(paste0("out",j), list(allreps, meandata))
}


par(mfcol=c(3,3), mar=c(4,4,0,0), oma=rep(1,4))
with(d1, plot(time, N, type='l', xlab="Time", ylab="Population size", main="c=0.05, bs=ds=0.01", ylim=c(5, 1.2*max(N))))
for(i in 1:40) with(subset(dat1, rep==i), lines(time, N, col=gray(0.8)))
with(d1, lines(time, N, lwd=2))

with(d1, plot(time, mean, type='l', xlab="Time", ylab="Mean bmax", ylim=c(1.5, 1.3*max(mean))))
for(i in 1:40) with(subset(dat1, rep==i), lines(time, mean, col=gray(0.8)))
with(d1, lines(time, mean, lwd=2))

with(d1, plot(time, var, type='l', xlab="Time", ylab="Var bmax", ylim=c(0, 1.5*max(var))))
for(i in 1:40) with(subset(dat1, rep==i), lines(time, var, col=gray(0.8)))
with(d1, lines(time, var, lwd=2))

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
